---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Build Status](https://travis-ci.org/avalcarcel9/rtapas.svg?branch=master)](https://travis-ci.org/avalcarcel9/rtapas)

[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/avalcarcel9/rtapas?branch=master&svg=true)](https://ci.appveyor.com/project/avalcarcel9/rtapas)

<!-- [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/rtapas)](https://cran.r-project.org/package=rtapas) -->

# rtapas

The goal of `rtapas` is to determine a subject-specific threshold to apply to multiple sclerosis lesion probability maps for automatic segmentation. This R package is based on the Thresholding Approach for Probability Map Automatic Segmentation method. The methods are in progress for publication. This package creates data structures necessary for training and testing the TAPAS model. The model can then be used to predict subject-specific threhsolds to use on probability maps for automatic lesion segmentation.

## Installation

This packages is still very much under construction and development. Please check back when the builds are stable.

To get the latest development version from GitHub:

```{r, eval = FALSE}
# install.packages('devtools')
devtools::install_github('avalcarcel9/mimosa')
```

We are currently working to get the package on [Neuroconductor](www.neuroconductor.org).

## Function Documentation

```{r, eval = FALSE}
rtapas::gen_tapas_data()
rtapas::train_tapas()
```


